Escape Analysis and Specialization in a Tracing JIT

### 2010-09-28 talk/pepm2011/paper.tex
cite psyco
swap last two arguments in setfield

The tracing JIT approach that the PyPy project is taking removes the overhead
of bytecode dispatch. In this paper we want to explain how the traces that are
produced by PyPy's tracing JIT can be optimized to also remove some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. To understand the problem more closely, we analyze the
occurring object lifetimes in Section~\ref{sec:lifetimes}. The most important
technique to achieve this is a form of escape analysis \cite{XXX} that we call
\emph{virtual objects},\footnote{The terminology comes from \cite{Psyco}}
which is described in Section~\ref{sec:virtuals}. The
goal of virtual objects is to remove allocations of temporary objects that have
a predictable lifetime and to optimize type dispatching in the process.

@@

\begin{figure}
\texttt{
\begin{tabular}{l} 
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = getfield($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = getfield($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{5}$, $i_{4}$, intval) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~setfield($p_{6}$, -100, intval) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = getfield($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = getfield($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{10}$, $i_{9}$, intval) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~setfield($p_{11}$, -1, intval) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = getfield($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = getfield($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{15}$, $i_{14}$, intval) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = getfield($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\label{fig:unopt-trace}
\caption{Unoptimized Trace for the Simple Object Model}
\end{figure}

@@

\texttt{
\begin{tabular}{l} 
$p_{5}$ = new(BoxedInteger) \\
setfield($p_{5}$, $i_{4}$, intval) \\
$p_{6}$ = new(BoxedInteger) \\
setfield($p_{6}$, -100, intval) \\
\end{tabular}
}

@@

\texttt{
\begin{tabular}{l} 
$p_{15}$ = new(BoxedInteger) \\
setfield($p_{15}$, $i_{14}$, intval) \\
$p_{10}$ = new(BoxedInteger) \\
setfield($p_{10}$, $i_{9}$, intval) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}

>>>>

The tracing JIT approach that the PyPy project is taking removes the overhead
of bytecode dispatch. In this paper we want to explain how the traces that are
produced by PyPy's tracing JIT can be optimized to also remove some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. To understand the problem more closely, we analyze the
occurring object lifetimes in Section~\ref{sec:lifetimes}. The most important
technique to achieve this is a form of escape analysis \cite{XXX} that we call
\emph{virtual objects}\footnote{The terminology comes from \cite{rigo_representation-based_2004}},
which is described in Section~\ref{sec:virtuals}. The
goal of virtual objects is to remove allocations of temporary objects that have
a predictable lifetime and to optimize type dispatching in the process.

@@

\begin{figure}
\texttt{
\begin{tabular}{l} 
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = getfield($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = getfield($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~setfield($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = getfield($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = getfield($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~setfield($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = getfield($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = getfield($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~setfield($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = getfield($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\label{fig:unopt-trace}
\caption{Unoptimized Trace for the Simple Object Model}
\end{figure}

@@

\texttt{
\begin{tabular}{l} 
$p_{5}$ = new(BoxedInteger) \\
setfield($p_{5}$, intval, $i_{4}$) \\
$p_{6}$ = new(BoxedInteger) \\
setfield($p_{6}$, intval, -100) \\
\end{tabular}
}

@@

\texttt{
\begin{tabular}{l} 
$p_{15}$ = new(BoxedInteger) \\
setfield($p_{15}$, intval, $i_{14}$) \\
$p_{10}$ = new(BoxedInteger) \\
setfield($p_{10}$, intval, $i_{9}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
